One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-11-12 paper.tex
add alt text to figures

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{enumerate}
\usepackage{caption}
\usepackage{textcomp}    % for Hawaii characters
\usepackage{url}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics{figures/study_area_design.pdf}
	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{}.  The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
	\label{fig:2002studyareapointspt}
\end{figure}

@@

\autoref{fig:intensity-mean-cv}A shows a map of the posterior mean intensity field alongside the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/intensity_mean_cv_sd.pdf}
		\caption{The A) mean, B) coefficient of variation and C) standard deviation of the posterior intensity field.  For the mean and standard deviaion the units are number of birds per m$^2$.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of higher intensity in the south and lower intensity in the north, agreeing with the broad patterns observed in a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity. The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.  Although survey effort is higher in the south, the increased uncertainty is likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.  For this reason we suggest using both the CV and SD to visualise uncertainty.

\autoref{fig:dist-mat-pairs} summarises the estimates of spatial structure in the model.  The posterior detection function (\autoref{fig:dist-mat-pairs}A) shows that the detection probability drops to around 0.25 at the maximum observable distance of 58 metres.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
		\caption{A) Posterior half-normal detection function. B) Posterior Mat\'ern correlation function for the GMRF spatial effect.  The black line is the posterior mean and the grey shaded area marks the 95\% credible interval. C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution. Red circles indicate the observed count of pairwise distances within each bin.}
		\label{fig:dist-mat-pairs}
	\end{center}
\end{figure}
The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B, this represents the degree of spatial smoothing in the posterior intensity field. We note that this correlation function shows positive correlation on the log-scale. On the intensity scale the empirical range of correlation is shorter.  \autoref{fig:dist-mat-pairs}C summarises the model's ability to replicate the spatial structure in the intensity and detection process simultaneously by comparing the frequency of pairwise distances between all observations and those on data simulated from the joint posterior of the fitted model.  This is based on 500 Monte Carlo samples and is essentially an adjusted version of Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which accounts for the distance sampling observation process.  The model shows a good ability to reconstruct the spatial structure due to the distance to observe and spatially varying intensity surface.

To visualise the posterior abundance for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.  Let $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$, the integral of the intensity field over the study region.  A point estimate can be calculated by integrating the posterior mean intensity. To generate a full posterior estimate for abundance we approximate the posterior as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$ where $\lambda^{(1)}, \ldots, \lambda^{(m)}$ are samples from the posterior intensity field. Due to the properties of the log-Gaussian Cox process, each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/N_posterior.pdf}
		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.  This approach can also be used to generate credible intervals and other posterior summaries of interest.

@@

Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although it is frustratingly common to see maps of point estimates without any accompanying communication of uncertainty.  The most common method to communicate uncertainty in maps of animal density is to produce maps of measures of predictive uncertainty, such as the standard deviation, CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}, posterior quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.  Whilst such maps are useful, they all mask certain properties of the intensity field that may be important when communicating the results of the analysis. For example, the posterior mean will always be smoother than realisations of the field.  If there is interest in understanding patterns of spatial clustering, the posterior mean map presents a misleadingly homogeneous picture.  This highlights the fact that even a relatively innocuous map of the posterior mean can mask certain features of the random field. \autoref{fig:intensity-realizations} shows three realisations of the posterior field, all of which show more heterogeneity in the high intensity region compared to the posterior mean.
\begin{figure}[!htb]
	\includegraphics{figures/intensity_realized.pdf}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}

@@

Quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.  Stakeholders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound. However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/intensity_quantiles.pdf}
		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}
These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the misleading nature of these maps, if we were to (incorrectly) treat the lower and upper quantile plots as though they were intensity surfaces and integrate them, then we would obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.  However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).  In other words, these maps tempt an interpretation that is inconsistent with the very model that produced the maps.

@@

\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).  Note that we use this threshold as the definition of a hotspot, and proceed with generating model results that incorporate this definition.  This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95. 
\begin{figure}[!htb]
	\includegraphics{figures/excursions.pdf}
	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  B) The positive excursion function with a level corresponding to 1 bird per hectare.  The scale shows the probability values $1-\alpha$.}
	\label{fig:excursions}
\end{figure} 
This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.  Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.  While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.  Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots).  To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.  This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.  That is, if we allow greater uncertainty, this function shows which locations would be included in the excursion set.

>>>>

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{enumerate}
\usepackage{caption}
\usepackage{textcomp}    % for Hawaii characters
\usepackage{url}
\usepackage{subcaption}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics{figures/study_area_design.pdf}
	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{}.  The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
	\subcaption*{Alt text: A map of the Hakalau study region with the locations of point transects.  The point transects are arranged approximately evenly spaced along straight lines moving roughly east to west.  The southern portion has higher survey effort comopared to the northern portion of the study area.}
	\label{fig:2002studyareapointspt}
\end{figure}

@@

\autoref{fig:intensity-mean-cv}A shows a map of the posterior mean intensity field alongside the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/intensity_mean_cv_sd.pdf}
		\caption{The A) mean, B) coefficient of variation and C) standard deviation of the posterior intensity field.  For the mean and standard deviaion the units are number of birds per m$^2$.}
		\subcaption*{Alt text:  A figure of 3 maps of the study region that summarise the posterior intensity field.  Map A shows the posterior mean showing a hot spot in the southern portion of the map. Map B shows the posterior coefficient of variation, showing higher values in the northern portion of the map. Map C shows the posterior standard deviation, showing higher standard deviation values in the southern portion of the map.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of higher intensity in the south and lower intensity in the north, agreeing with the broad patterns observed in a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity. The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.  Although survey effort is higher in the south, the increased uncertainty is likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.  For this reason we suggest using both the CV and SD to visualise uncertainty.

\autoref{fig:dist-mat-pairs} summarises the estimates of spatial structure in the model.  The posterior detection function (\autoref{fig:dist-mat-pairs}A) shows that the detection probability drops to around 0.25 at the maximum observable distance of 58 metres.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
		\caption{A) Posterior half-normal detection function. B) Posterior Mat\'ern correlation function for the GMRF spatial effect.  The black line is the posterior mean and the grey shaded area marks the 95\% credible interval. C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution. Red circles indicate the observed count of pairwise distances within each bin.}
		\subcaption*{Alt text:  A figure summarising the estimated spatial structure in the point transect distance sampling data.  Sub-figure A shows the posterior detection function with associated uncertainty.  Detectability drops to just above 0.25 at the maximum observable distance.  Sub-figure B shows the posterior Mat\`ern correlation function for the random effect.  It shows a decreasing correlation as distance increases, dropping to a correlation around 0.1 at a distance of 10 kilometres.  Sub-figure C shows boxplots that summarise the posterior distribution of pairwise distances between observations simulated from the posterior predictive distribution.  It shows a very close match between the posterior mean and pairwise distances observed in the real observations.}
		\label{fig:dist-mat-pairs}
	\end{center}
\end{figure}
The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B, this represents the degree of spatial smoothing in the posterior intensity field. We note that this correlation function shows positive correlation on the log-scale. On the intensity scale the empirical range of correlation is shorter.  \autoref{fig:dist-mat-pairs}C summarises the model's ability to replicate the spatial structure in the intensity and detection process simultaneously by comparing the frequency of pairwise distances between all observations and those on data simulated from the joint posterior of the fitted model.  This is based on 500 Monte Carlo samples and is essentially an adjusted version of Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which accounts for the distance sampling observation process.  The model shows a good ability to reconstruct the spatial structure due to the distance to observe and spatially varying intensity surface.

To visualise the posterior abundance for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.  Let $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$, the integral of the intensity field over the study region.  A point estimate can be calculated by integrating the posterior mean intensity. To generate a full posterior estimate for abundance we approximate the posterior as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$ where $\lambda^{(1)}, \ldots, \lambda^{(m)}$ are samples from the posterior intensity field. Due to the properties of the log-Gaussian Cox process, each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/N_posterior.pdf}
		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
		\subcaption*{Alt text:  A figure summarising the posterior for abundance.  The posterior is approximately bell-shaped with the bulk of the posterior mass between 4000 and 8000 birds.}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.  This approach can also be used to generate credible intervals and other posterior summaries of interest.

@@

Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although it is frustratingly common to see maps of point estimates without any accompanying communication of uncertainty.  The most common method to communicate uncertainty in maps of animal density is to produce maps of measures of predictive uncertainty, such as the standard deviation, CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}, posterior quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.  Whilst such maps are useful, they all mask certain properties of the intensity field that may be important when communicating the results of the analysis. For example, the posterior mean will always be smoother than realisations of the field.  If there is interest in understanding patterns of spatial clustering, the posterior mean map presents a misleadingly homogeneous picture.  This highlights the fact that even a relatively innocuous map of the posterior mean can mask certain features of the random field. \autoref{fig:intensity-realizations} shows three realisations of the posterior field, all of which show more heterogeneity in the high intensity region compared to the posterior mean.
\begin{figure}[!htb]
	\includegraphics{figures/intensity_realized.pdf}
	\caption{Three realizations of the posterior intensity field}
	\subcaption*{Alt text: A figure showing three realisations of the posterior intensity field.  The realisations all show broadly the same trend as the posterior mean but with more variability and a shorter effective range of correlation.}
	\label{fig:intensity-realizations}
\end{figure}

@@

Quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.  Stakeholders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound. However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/intensity_quantiles.pdf}
		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
		\subcaption*{Alt text: 2 maps summarising pointwise prediction quantiles for the posterior intensity field. Map A shows the 0.025 quantile, and map B shows the 0.975 quantile.}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}
These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the misleading nature of these maps, if we were to (incorrectly) treat the lower and upper quantile plots as though they were intensity surfaces and integrate them, then we would obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.  However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).  In other words, these maps tempt an interpretation that is inconsistent with the very model that produced the maps.

@@

\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).  Note that we use this threshold as the definition of a hotspot, and proceed with generating model results that incorporate this definition.  This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95. 
\begin{figure}[!htb]
	\includegraphics{figures/excursions.pdf}
	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  B) The positive excursion function with a level corresponding to 1 bird per hectare.  The scale shows the probability values $1-\alpha$.}
	\subcaption*{Alt text: Two maps showing estimates of excursion set and excursion functions.  Map A shows the excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  The map shows a region in the south and a very small region slightly north of this that are within the excusion set.  Map B shows the excursion function, showing that most of the hotspot would remain in the excursion set even under probability values very close to 1.  A few areas additional areas would be included if the probability were lower than 0.95 but this are on the fringe and are not substantially larger excursion sets than the 0.95 example.}
	\label{fig:excursions}
\end{figure} 
This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.  Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.  While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.  Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots).  To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.  This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.  That is, if we allow greater uncertainty, this function shows which locations would be included in the excursion set.
